An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess

### 2022-12-10 Manuscript/manuscript.tex
fixed citations

Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen,  and “alleged cheating” in the world of professional chess. The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round. This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers.

@@

Chess.com, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.

Of these cheat detection methods, the particular focus of this paper will be in investigating item 2. Item 2 requires an assessment of move quality which is usually done in centipawns. The centipawn is a unit of measure used in chess as a representation of advantage. One centipawn is equal to one hundredth of a pawn and, although its value plays no formal role in the game, they are used by chess engines and players alike as a means to evaluate the positions of a chess board via centipawn loss. Centipawn loss (CPL) is a calculation and numerical score given by a chess engine as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine. Thus, average centipawn loss (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly speaking, an average assessment of move quality as computed by an engine. In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been. Although the details of how centipawn loss is calculated will not be explicitly covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory in chess, can be introduced thusly.

Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player \begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and \begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} actual move in centipawns. Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math}.

Since the specific details of the methods that Chess.com uses in assessing move quality have not been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified by ELO score, could feasibly serve as the best way to replicate how they would compare a players past performance to their historical strengths. With respect to the Niemann controversy, this hypothesized method has been used by many as an attempt to prove or disprove whether he had “cheated” in the past. This paper will attempt to investigate whether the use of ACPL as a means to verify the integrity of chess players in the age of modern computing is an appropriate method or just a commonly peddled misconception using data from Hans Niemann's past games.

@@

The dataset was originally compiled by Rafael Leite, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation. This data itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games. Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis. He also expanded the data to include the ACPL of each player in each game as computed by the chess engine, Stockfish.

@@

The purpose of this research was meant to analyze how one could detect foul play in chess by observing the correlation between a player's rating and ACPL over a period of time; but rather than painting a clearer picture of what many consider to be a common method to identify cheaters, more questions have been raised than answers. In the case of Hans Niemann, evidently the data does not even suggest a strong linear relationship between his ELO score and ACPL over time. Upon further research, it appears that the claims made by many regarding the existence of this relationship is somewhat contested. Regrettably, the only real contribution this paper provides is that, in the case of Hans Niemann's professional history, it is inconclusive whether the methodologies used here can even be used to accurately indicate “cheating”. The fact that the data used in this paper only reflects the history of a single professional player may be a significant limitation in investigating the true relationship of these variables. Shifting the focus from Niemann and performing the same analysis on a dataset consisting of players that were randomly selected from another database may provide more accurate results of what one could expect this relationship could be for the average player. Another definite limitation of this study came about as a result during the process of binning observations before plotting the central tendency of ELO score against ACPL. It is possible that during this process, the visibility of certain observations that could have been suspicious would undoubtedly be lessened when binned with others that were less so. This would have effectively drowned out any times where Niemann could possibly have “cheated” inconsistently. Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating. Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater. A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested. This would only give them a slight advantage, but it would be nearly impossible to detect using a simple correlation. The possible future directions of this study at the moment would be to investigate this same relationship between ELO score and ACPL for other professional chess players to observe if the same or different effects could be observed.

>>>>

Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen,  and “alleged cheating” in the world of professional chess. The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round. This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers\citep{wsj2022}.

@@

Chess.com\citet{chess2022cheating}, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.

Of these cheat detection methods, the particular focus of this paper will be in investigating item 2. Item 2 requires an assessment of move quality which is usually done in centipawns. The centipawn is a unit of measure used in chess as a representation of advantage. One centipawn is equal to one hundredth of a pawn and, although its value plays no formal role in the game, they are used by chess engines and players alike as a means to evaluate the positions of a chess board via centipawn loss. Centipawn loss (CPL) is a calculation and numerical score given by a chess engine as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine. Thus, average centipawn loss (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly speaking, an average assessment of move quality as computed by an engine. In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been\citep{avva2022guess}. Although the details of how centipawn loss is calculated will not be explicitly covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory in chess, can be introduced thusly.

Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player \begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and \begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} actual move in centipawns. Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math}\citet{anbarciai}.

Since the specific details of the methods that Chess.com uses in assessing move quality have not been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified by ELO score, could feasibly serve as the best way to replicate how they would compare a players past performance to their historical strengths. With respect to the Niemann controversy, this hypothesized method has been used by many\citet{medium2022} as an attempt to prove or disprove whether he had “cheated” in the past. This paper will attempt to investigate whether the use of ACPL as a means to verify the integrity of chess players in the age of modern computing is an appropriate method or just a commonly peddled misconception using data from Hans Niemann's past games.

@@

The dataset was originally compiled by Rafael Leite\citep{leite2022}, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation. This data\citep{chessbase} itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games. Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis. He also expanded the data to include the ACPL of each player in each game as computed by the chess engine, Stockfish.

@@

The purpose of this research was meant to analyze how one could detect foul play in chess by observing the correlation between a player's rating and ACPL over a period of time; but rather than painting a clearer picture of what many consider to be a common method to identify cheaters, more questions have been raised than answers. In the case of Hans Niemann, evidently the data does not even suggest a strong linear relationship between his ELO score and ACPL over time. Upon further research, it appears that the claims made by many regarding the existence of this relationship is somewhat contested. Regrettably, the only real contribution this paper provides is that, in the case of Hans Niemann's professional history, it is inconclusive whether the methodologies used here can even be used to accurately indicate “cheating”. The fact that the data used in this paper only reflects the history of a single professional player may be a significant limitation in investigating the true relationship of these variables. Shifting the focus from Niemann and performing the same analysis on a dataset consisting of players that were randomly selected from another database may provide more accurate results of what one could expect this relationship could be for the average player. Another definite limitation of this study came about as a result during the process of binning observations before plotting the central tendency of ELO score against ACPL. It is possible that during this process, the visibility of certain observations that could have been suspicious would undoubtedly be lessened when binned with others that were less so. This would have effectively drowned out any times where Niemann could possibly have “cheated” inconsistently. Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating. Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater\citep{chessbot}. A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested. This would only give them a slight advantage, but it would be nearly impossible to detect using a simple correlation. The possible future directions of this study at the moment would be to investigate this same relationship between ELO score and ACPL for other professional chess players to observe if the same or different effects could be observed.
